---
title: "Extra Credit DESeq"
author: "Divyanshu Kawankar"
date: "11/22/2021"
output:
  pdf_document: default
  html_document: default
---

#Section 1. Differential Expression Analysis

```{r}

library(DESeq2)

#load data files
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

```

```{r}
#import metadata and take a look
colData = read.csv(metaFile, row.names=1)
head(colData)

#import countdata and take a look
countData = read.csv(countFile, row.names=1)
head(countData)
```
> Q. Complete the code below to remove the troublesome first column from countData
  - Refer to code below 

```{r}

countData <- as.matrix(countData[,-1])
head(countData)

```

>Q. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns).
 - Refer to below

```{r}
countData = countData[rowSums(countData != 0) > 0,]

head(countData)
```

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
dds
```

```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
```


```{r}
summary(res)
```

#Volcono plot

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

>Q. Improve this plot by completing the below code, which adds color and axis labels
  -refer to code below

```{r}
mycols <- rep("gray", nrow(res) )

mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

Adding Gene Notation

>Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below. 
  - Refer to code below

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```

> Q. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory.
  -Refer to code below
  
```{r}
res = res[order(res$pvalue),]
write.csv(res,"deseq_results.csv")
```

#Pathway Analysis

```{r}
library(pathview)

library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)

attributes(keggres)

```

```{r}
# Look at the first few down (less) pathways
head(keggres$less)

pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

```{r}
# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

```{r}
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

```{r}
#Draw plots for top 5 pathways

pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

> Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-reguled pathways?
  - Yes. If you use tail instead of Head
  
#Gene Ontology (GO)

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

#Reactome Analysis

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

> Q: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?
  - The endousomal/Vascular Pathway. They do not match the previous pathways. Differences between the two factors could be different databases. We used the Gene ontology pathway instead which could have been the cause of the difference.